Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-16 paper.tex
More systematic treatment of data files.

\begin{figure}
% \begin{center}
% \begin{tabular}{cc}
% % \includegraphics[width=5cm]{msprime-scaling.png} &
% % \includegraphics[width=5cm]{figures/recombination} \\
% \end{tabular}
% \end{center}
\includegraphics[width=\textwidth]{figures/recombination}
\caption{\label{fig-recombination-perf}(A) Approximate run time for \msprime\
simulations as a function of scaled recombination rate.
(B) Comparison of the ARG simulations vs minimal tree sequence representation.}
\end{figure}

@@

For our purposes, an ARG is a realisation of the coalescent with
recombination, in the Griffiths (little ARG) sense.
As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
works by dynamically generating and traversing a little ARG.
The graph is not explicitly represented in memory, but is partially
present through the extant lineages and the ancestral material they carry
over time. Since the ARG is a large and unwieldy object
(Fig.~\ref{fig-recombination-perf}B) we do not use it as output, but
rather store the information required to recover the genealogical
history as nodes and edges in a tree sequence (see section XX),
or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.

Although the tree sequences output by \msprime\ contain full information
about the genealogical trees, their correlation structure along the genome,
and the ancestral haplotypes on which coalescences occured, some information
is lost in this mapping down from ARG space to the minimal tree sequence
form. In particular, we lose
information about ancestral haplotypes that were common ancestors but
in which no coalescences occured, and also information about the precise time
and genomic location of recombination events. In most cases, such
information is of little relevance as it is in principle unknowable,
but there are occasions such as visualisation or computing likelihoods (see
below) in which is useful. We therefore provide
the \texttt{record\_full\_arg} option in \msprime\
to store a representation of the complete ARG traversed during simulation.
This is done by storing extra nodes (marked with specific flags, so they
can be easily identified later) and edges in the tree sequence.
Fig.~\ref{fig-recombination-perf}B shows that this extra ARG information
quickly comes to dominate the output as we increase the sequence length,
and the resulting tree sequence files are many times larger
than the standard minimal tree sequence representation.

@@

>>>>

\begin{figure}
% \begin{center}
% \begin{tabular}{cc}
% % \includegraphics[width=5cm]{msprime-scaling.png} &
% % \includegraphics[width=5cm]{figures/recombination} \\
% \end{tabular}
% \end{center}
% \includegraphics[width=\textwidth]{figures/recombination}
\Large{Plot of msprime run time against $\rho$}
\caption{\label{fig-recombination-perf} Approximate run time for \msprime\
simulations as a function of scaled recombination rate.
}
\end{figure}

@@

@@

\begin{figure}
\begin{center}
\includegraphics[width=7cm]{figures/arg}
\end{center}
\caption{\label{fig-arg} A figure illustrating the relative size of
the full ARG representation compared the minimal tree sequence.
\jkcomment{This figure is quite preliminary and we need to improve it.}}
\end{figure}

For our purposes, an ARG is a realisation of the coalescent with
recombination, in the Griffiths (little ARG) sense.
As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
works by dynamically generating and traversing a little ARG.
The graph is not explicitly represented in memory, but is partially
present through the extant lineages and the ancestral material they carry
over time. Since the ARG is a large and unwieldy object
(Fig.~\ref{fig-arg}) we do not use it as output, but
rather store the information required to recover the genealogical
history as nodes and edges in a tree sequence (see section XX),
or ``coalescence records'' in the terminology of~\citep{kelleher2016efficient}.

Although the tree sequences output by \msprime\ contain full information
about the genealogical trees, their correlation structure along the genome,
and the ancestral haplotypes on which coalescences occured, some information
is lost in this mapping down from ARG space to the minimal tree sequence
form. In particular, we lose
information about ancestral haplotypes that were common ancestors but
in which no coalescences occured, and also information about the precise time
and genomic location of recombination events. In most cases, such
information is of little relevance as it is in principle unknowable,
but there are occasions such as visualisation or computing likelihoods (see
below) in which is useful. We therefore provide
the \texttt{record\_full\_arg} option in \msprime\
to store a representation of the complete ARG traversed during simulation.
This is done by storing extra nodes (marked with specific flags, so they
can be easily identified later) and edges in the tree sequence.
Fig.~\ref{fig-arg} shows that this extra ARG information
quickly comes to dominate the output as we increase the sequence length,
and the resulting tree sequence files are many times larger
than the standard minimal tree sequence representation.
